---
title: "Final Project"
author: "Shanie Lynn Jorgenson"
date: "2025-11-18"
output: html_document
---

Introduction:

Background & literature foreshadow the importance of each question; clear statement of 2+ questions addressing a new dimension of bee conservation, ecology, or management using OBA or student’s data. Graduate students write as a thesis‑style intro with comprehensive background literature. Citations use a standardized style. ~1 page.

1. Data: (1) List 2+ datasets and 3+ distinct data skills (e.g., table manipulation, interaction data, raster, vector). Provide general metadata (as in workflow). (2) Read and tidy 2018–2024 OBA (plant taxonomy cleaned via regex if using plant data). If using own data, it is complete and cleaned.

```{r}

# Import Data Sets:

# OBA Data:
oba <- read.csv("data/OBA_2018-2024.csv")

# Land/vegitation Data:

```

```{r}

# Data wrangling, summarizing, exploring: Relevant manipulation & summaries; exploratory tables and plots (scatter, hist, bar, etc.); multiple relevant visualizations and data checks.

```

```{r}

# Hypothesis test: Use a simulation or statistical test to make an inference; grad students use the best available statistical approach.	

```

```{r}

# Visualization: Final, publication‑quality figures that answer the questions; clear labels, legends, and units.

```

```{r}

# Additional tool/analysis: Extra credit: grads must incorporate a tool not addressed in class in analysis/visualization/workflow; undergrads may earn EC.

```

Conclusion: Clearly answers the original questions and connects to relevant comparative literature.

